---
title: "importance_sampling"
author: "zihao12"
date: "2020-12-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
set.seed(123)
```


# Problem 2
Let $X \sim t_3$. Want to compute $\alpha = P(X > 1000)$. Use importance sampling.

Theoretical result is:
```{r}
pt(q = 1000, df = 3, lower.tail = FALSE)
```
If we are to use traditional Monte Carlo, which is $\frac{1}{n} \sum_{i=1}^n 1_{[X > 1000]}$, we need a lot of samples to get any $X > 1000$. 

## Importance Sampling
I use Cauchy as it is heavy-tailed. The formula is $$E_p[f(X)] = \int f(x) p(x) dx = \int f(x) \frac{p(x)}{q(x)} q(x) dx = E_q[f(x) \frac{p(x)}{q(x)}] \approx \frac{1}{n} \sum_{i=1}^n f(x_i) \frac{p(x_i)}{q(x_i)}$$

SO we can easily get the estimate below. 
```{r}
n = 1e6
x = rcauchy(n = n, location = 0, scale = 1)
w = exp(  dt(x = x, df = 3, log = TRUE) - dcauchy(x = x, location = 0, scale = 1, log = TRUE))
mean((x > 1000) * w)
```


## Compare with Monte Carlo method. 
I tried with $10^8$ samples, but still can't get anything above $1000$. 
```{r}
# mean(rt(n = 1e4, df = 3) > 1000)
# mean(rt(n = 1e6, df = 3) > 1000)
mean(rt(n = 1e8, df = 3) > 1000)
```
So if we want to achieve the same level of accuarcy of Importance Sampling with Cauchy (around $10^{-11}$), we need more than $10^8$ samples, way more than $10^6$ samples used in Importance Sampling.

# Problem 1
Use Rejection Sampling to sample from Mises distribution. Since both $f, q$ (pdf for objective and propsal distributions) take support in $[-\pi, \pi]$, and $f(x) = C e^{K cos(x)} \leq C e^K$, $q(x) = \frac{1}{2\pi}$, we can choose $M = \frac{C e^K}{2 \pi}$. This gives us acceptance probability of $\frac{e^K cos(x)}{e^K}$ for each $x$. Turned into code below:
```{r}
rejection_sampler <- function(k, n = 1e3){
  x = runif(n = n, min = -pi, max = pi)  ## proposals
  accept_prob = exp(k * cos(x))/exp(k)
  accept = rbinom(n = n, size = 1, prob = accept_prob)
  return(x[accept == 1])
}

dmises <- function(x, k){
  const = 2*pi*besselI(x = k, nu = 0)
  return(exp(k*cos(x))/const)
}
```


```{r}
## try when k = 1
k = 1
x = rejection_sampler(k = k, n = 1e6)
hist(x, probability = TRUE, breaks = 100, main = "sampled vs theoretical (kappa = 1)")
lines(sort(x), dmises(x = sort(x), k = k), col = "red")

## try when k = 100
k = 100
x = rejection_sampler(k = k, n = 1e6)
hist(x, probability = TRUE, breaks = 100, main = "sampled vs theoretical (kappa = 100)")
lines(sort(x), dmises(x = sort(x), k = k), col = "red")
```



