---
title: "sparseMatrix"
author: "zihao12"
date: "2020-05-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

* I want to learn about `sparseMatrix` in R.\

* denote our matrix $X$, with `dim(X) = c(n, p)` and number of nnzero elements `d`. 


```{r}
library(Matrix)
```


## `dgTMatrix`
The matrix is stored in triplet $(i, j, x_{ij})$. 

### basics
```{r}
### create a dgTMatrix
RowIdx = c(1,1,2,3,3,4)
ColIdx = c(2,2,7,7,9,9)
## this creates a TRUE/FALSE matrix
X <- sparseMatrix(
  i = RowIdx,
  j = ColIdx,
  dims = c(5, 10)
)
summary(X)

X
## let each x_ij = 1 (note (1,2) appears twice)
X <- sparseMatrix(
  i = RowIdx,
  j = ColIdx,
  dims = c(5, 10),
  x = 1L ## or a vector of the length as RowIdx
)
summary(X)
X

## see what does X have
## note that row and col index starts from 0!
str(X)
```

### efficient functions over elements
[Peter](https://pcarbo.github.io/) provides a good example of efficient function

```{r}
# Apply operation f to all nonzeros of a sparse matrix.
apply.nonzeros <- function (X, f) {
  d <- summary(X)
  return(sparseMatrix(i = d$i,j = d$j,x = f(d$x),dims = dim(X)))
}
```


## `dgCMatrix`

```{r}
Y <- as(X, "dgCMatrix")
Y
str(Y)
```

*  `i` is the row index\

* `p[k], k >= 1` is the number of nonzero data points counting up to column `k-1`. We can get the number of nonzero elements in each column this way:
```{r}
diff(Y@p)
```
We can get to `j` from `p` and `i` as well:
```{r}
rep(1:Y@Dim[2], diff(Y@p))
## another way to do it
summary(Y)$j
## compare with true ColIdx
ColIdx
```

## compare size
```{r}
# A large matrix
set.seed(1)
## use the default storage class
m <- sparseMatrix(
  i = sample(x = 1e4, size = 1e4),
  j = sample(x = 1e4, size = 1e4),
  #x = rnorm(n = 1e4)
  x = rpois(n = 1e4, lambda = 100) 
)
class(m)
dim(m)

pryr::object_size(m)

pryr::object_size(as(m, "dgTMatrix"))

pryr::object_size(as.matrix(m))

```

## copare computation speed
```{r}
m2 <- sparseMatrix(
  i = sample(x = 1e4, size = 1e4),
  j = sample(x = 1e4, size = 1e4),
  #x = rnorm(n = 1e4)
  x = rpois(n = 1e4, lambda = 100) 
)

system.time(
  test <- replicate(100, m %*% m2)
)

m_ = as(m, "dgTMatrix")
m2_ = as(m2, "dgTMatrix")
system.time(
  test <- replicate(100, m_ %*% m2_)
)

## takes too long to run
# system.time(
#   test <- as.matrix(m) %*% t(as.matrix(m))
# )
```

how to explain the performance difference and when to use which?

does summary take time?
```{r}
class(m)
system.time(
  test <- replicate(100, summary(m))
)

class(m_)
system.time(
  test <- replicate(100, summary(m_))
)

```

